Optimal updating factor in adaptive flat-distribution simulations

### 2016-04-22 doc/invt.tex
invt.tex: move Gamma values to Appendix

In the asymptotic regime
(i.e., if the simulation length, $T$, is sufficiently long),
we can approximate $\kappa_k(t)$ as
%
\begin{equation}
  \kappa_k(t) \approx \Gamma_k \, \delta(t)
  ,
\label{eq:kappa_delta}
\end{equation}
%
where
\begin{equation}
  \Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt,
  \notag
  %\label{eq:Gamma_integral}
\end{equation}
or in discrete time,
\begin{equation}
  \Gamma_k = \sum_{t = -\infty}^\infty \kappa_k(t).
  \label{eq:Gamma_sum}
\end{equation}
%
Then, Eq. \eqref{eq:error_asym} is simplified as
%
\begin{align}
  \Err_A
  &=
  \int_0^T
  \sum_{k = 2}^n
  \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr) \, dt.
\label{eq:error_asym1}
\end{align}
%
In this way,
the underlying sampling process
affects the error only through the set of numbers, $\Gamma_i$'s.

\subsubsection{\label{sec:Gamma}
Integrals of the autocorrelation functions}

We can evaluate the integrals of the autocorrelation functions,
$\Gamma_k$'s,
in two special cases.
%
First, if the sampling is perfect,
Eq. \eqref{eq:kappa_delta} is exact, and
%
\begin{equation}
  \Gamma_k = 1.
\label{eq:Gamma_perfect}
\end{equation}
%
To see this, we note that
the bin index $i$ is an independent random variable
at each time step $t$ in this case, and
%
\begin{align}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  &=
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle
  \langle h_j(t') \rangle
  \notag
  \\
  &=
  \left( p_i \, \delta_{ij} - p_i \, p_j \right) \,
  \delta_{t, t'}
  ,
\label{eq:zz_perfect}
\end{align}
%
where $\delta(t - t')$ is the equivalent of $\delta_{t, t'}$
in the continuous limit.
%
\note{To show Eq. \eqref{eq:zz_perfect}, we first note that
  with $t \ne t'$ (going back to the discrete time case),
  $$
  \begin{aligned}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  &=
  \left\langle
    \zeta_i(t)
  \right\rangle
  \left\langle
    \zeta_j(t')
  \right\rangle
  \\
  &=
  \bigl\langle h_i(t)  - \langle h_i(t)  \rangle \bigr\rangle
  \bigl\langle h_j(t') - \langle h_j(t') \rangle \bigr\rangle
  =
  0.
  \end{aligned}
  $$
  Next, for $t = t'$, we have
  $$
  \zeta_i(t)
  = h_i(t) - \langle h_i(t) \rangle
  = \delta_{i, i(t)} - p_i,
  $$
  and
  $$
  \begin{aligned}
  \left\langle
    \zeta_i(t) \, \zeta_j(t)
  \right\rangle
  &=
  \left\langle
    \delta_{i, i(t)} \, \delta_{j, i(t)}
  \right\rangle
  -
  p_i \, p_j
  \\
  &=
  \left\langle
    \delta_{i, i(t)}
  \right\rangle
  \delta_{i, j}
  -
  p_i \, p_j
  =
  p_i \, \delta_{ij}
  - p_i \, p_j.
  \end{aligned}
  $$
  %
  The average
  $\left\langle \delta_{i, i(t)} \right\rangle$ is $p_i$,
  because as $i(t)$ goes through $1$ to $n$,
  $\delta_{i, i(t)}$ gives $1$
  if $i(t)$ matches $i$ with probability $p_i$,
  or $0$ otherwise.
}
%
Thus,
for $k \ge 2$, we get
%
\begin{equation}
  \kappa_k(t - t')
  =
  \sum_{i,j = 1}^n
  \frac{ \phi_{ki} } { p_i }
  \frac{ \phi_{kj} } { p_j }
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  %
  = \delta_{t, t'},
\notag
%\label{eq:kappa_perfect}
\end{equation}
%
where we have used
Eqs. \eqref{eq:eig_orthonormal_rows}-\eqref{eq:ortho1},
\eqref{eq:eta_def},
and
\eqref{eq:zz_perfect}.
%
\note{Proof.%Eq. \eqref{eq:kappa_perfect},
  $$
  \begin{aligned}
  \kappa_k(t - t')
  &=
  \sum_{i,j = 1}^n
  \frac{ \phi_{ki} } { p_i }
  \frac{ \phi_{kj} } { p_j }
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  \\
  %
  &=
  \sum_{i,j = 1}^n
  \frac{ \phi_{ki} } { p_i }
  \frac{ \phi_{kj} } { p_j }
  \left(
    p_i \, \delta_{i, j} - p_i \, p_j
  \right)
  \, \delta_{t, t'}
  \\
  &=
  \left(
    \sum_{j = 1}^n
    \frac{ \phi_{kj}^2 } { p_j }
    -
    \sum_{i = 1}^n \phi_{ki}
    \sum_{j = 1}^n \phi_{kj}
  \right)
  \, \delta_{t, t'}
  \\
  &=
  (1 - 0 \cdot 0) \, \delta_{t, t'}.
  \end{aligned}
  $$
  where,
  we have used
  Eq. \eqref{eq:eta_def}
  on the first line,
  %
  Eq. \eqref{eq:zz_perfect}
  on the second line,
  and
  Eq. \eqref{eq:eig_orthonormal_rows}
  on the last line.
}

Another special case is the
one-dimensional one-step process\cite{vankampen},
which models a local sampling process.
In each step, the system hops to
one of the two neighboring bins
with equal probability $1/2$.
%
The transition matrix is
%
\begin{equation}
  \arraycolsep=3.6pt\def\arraystretch{1.4}
  \mathbf A
  =
  \left(
    \begin{array}{cccccccc}
      \frac \sigma 2 & \frac 1 2 & 0 & \dots & \frac {1 - \sigma} 2 \\
      \frac 1 2 & 0         & \frac 1 2 & \dots & 0 \\
      \vdots & &  & & \vdots \\
      \frac {1 - \sigma} 2 & \dots &  & \frac 1 2 & \frac \sigma 2
    \end{array}
  \right),
\notag
%\label{eq:T_nn}
\end{equation}
%
where $\sigma = 0$ for a periodic variable, or $1$ otherwise.
%
For many updating schemes
(cf. Sec. \ref{sec:band-matrix}),
we can further assume that
the matrices $\mathbf A$ and $\mathbf w$
share the eigenvectors, but not the eigenvalues.
%
The eigenvalues of $\mathbf A$ are
$$
\Lambda_k
=
1 -
2 \sin^2 \frac{ (k - 1) \, g \, \pi }
              {       2 \, n        }
,
$$
where
$g = 2$ for a periodic variable, or $1$ otherwise.
%Note that these $\Lambda_k$'s are
%not to be confused with the eigenvalues of $\mathbf w$,
%$\lambda_k$'s.
%
In this case,
the autocorrelation function $\kappa_k(t)$
decays as $\kappa_k(0) \, \Lambda_k^t$,
with $\kappa_k(0) = 1$ being the same as
the value in perfect sampling.
%
Thus,
%$\Gamma_k$ is roughly twice the autocorrelation time, and
from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
  \Gamma_k
  =
  1 + 2 \, \sum_{t = 1}^\infty \Lambda_k^t
  =
  \frac{ 1 + \Lambda_k }
       { 1 - \Lambda_k }
  =
  \cot^2 \left[
    \frac{ (k - 1) \, g \, \pi }
         { 2 \, n }
  \right]
  .
\label{eq:Gamma_onestep}
\end{equation}
%

@@

@@

Here we give a program of computing
the optimal schedule and the error
of a given adaptive FDS simulation.
%
Basically, we need to extract
$\lambda_i$'s and $\Gamma_i$'s
from the input,
and then feed them to Eq. \eqref{eq:q_opt}
for the schedule.
%
The dependence of variables are shown in Fig. \ref{fig:vardep}.

@@

\subsection{\label{sec:measure_gamma}
Measurement of the integrals
of the autocorrelation functions}

%To use Eq. \eqref{eq:q_opt} in a real simulation,
%we need to know both $\lambda_k$'s and $\Gamma_k$'s.
%
While the updating scheme alone
determines the eigenmodes and $\lambda_k$'s,
the values of $\Gamma_k$'s depend also on the system
and the underlying sampling process (MC or MD).
%
%Below is a technique of measuring $\Gamma_k$'s.
%
%The idea is to run
To measure $\Gamma_k$'s, we can run
a simulation with a constant schedule,
$\alpha(t) = \alpha_0$,
and then use Eq. \eqref{eq:y2_eql}.
%
%A practical difficulty is that
However,
we often do not know the intrinsic distribution, $p^*_i$,
in advance to compute
the shifted bias potential, $v_i$,
from the original one, $V_i$
(cf. Sec. \ref{sec:FDS}).
%
To circumvent the problem, we
modify Eq. \eqref{eq:y2_eql} as
%
\begin{equation}
  \operatorname{var} Y_k
  =
  \frac{1}{2}
  \alpha \, \Gamma_k \, \lambda_k,
\label{eq:varY}
\end{equation}
%
where
``$\operatorname{var}$''
means the variance from time averaging,
and
%
\begin{align}
Y_k
&=
\sum_{ i = 1 }^n
  \phi_{k i} \, (V_i - \bar V),
\label{eq:Y_def}
\end{align}
%
with
$
\bar V
\equiv
\sum_{ j = 1 }^n p_j \, V_j.
$

\subsection{\label{sec:program_list}
Program
}

The complete program is the following.

>>>>

In the asymptotic regime
(i.e., if the simulation length, $T$, is sufficiently long),
we can approximate $\kappa_k(t)$ as
%
\begin{equation}
  \kappa_k(t) \approx \Gamma_k \, \delta(t)
  ,
\label{eq:kappa_delta}
\end{equation}
%
where
\begin{equation}
  \Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt,
  \notag
  %\label{eq:Gamma_integral}
\end{equation}
or in discrete time,
\begin{equation}
  \Gamma_k = \sum_{t = -\infty}^\infty \kappa_k(t).
  \label{eq:Gamma_sum}
\end{equation}
%
Then, Eq. \eqref{eq:error_asym} is simplified as
%
\begin{align}
  \Err_A
  &=
  \int_0^T
  \sum_{k = 2}^n
  \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr) \, dt.
\label{eq:error_asym1}
\end{align}
%
In this way,
the underlying sampling process
affects the error only through the set of numbers,
$\Gamma_k$'s.
%
In Appendix \ref{sec:Gamma},
we evaluate $\Gamma_k$'s in two special cases,
and show how to measure them
from an adaptive FDS simulation.

@@

\section{\label{sec:Gamma}
Integrals of the autocorrelation functions}

Here, we evaluate the integrals of
the autocorrelation functions,
$\Gamma_k$'s,
in two special cases
in Secs. \ref{sec:Gamma_perfect}
and \ref{sec:Gamma_onestep},
and describe a general method of measuring them
in Sec. \ref{sec:Gamma_measure}.

\subsection{\label{sec:Gamma_perfect}
Perfect sampling}

First, if the sampling is perfect,
Eq. \eqref{eq:kappa_delta} is exact, and
%
\begin{equation}
  \Gamma_k = 1.
\label{eq:Gamma_perfect}
\end{equation}
%
To see this, we note that
the bin index $i$ is an independent random variable
at each time step $t$ in this case, and
%
\begin{align}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  &=
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle
  \langle h_j(t') \rangle
  \notag
  \\
  &=
  \left( p_i \, \delta_{ij} - p_i \, p_j \right) \,
  \delta_{t, t'}
  ,
\label{eq:zz_perfect}
\end{align}
%
where $\delta(t - t')$ is the equivalent of $\delta_{t, t'}$
in the continuous limit.
%
\note{To show Eq. \eqref{eq:zz_perfect}, we first note that
  with $t \ne t'$ (going back to the discrete time case),
  $$
  \begin{aligned}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  &=
  \left\langle
    \zeta_i(t)
  \right\rangle
  \left\langle
    \zeta_j(t')
  \right\rangle
  \\
  &=
  \bigl\langle h_i(t)  - \langle h_i(t)  \rangle \bigr\rangle
  \bigl\langle h_j(t') - \langle h_j(t') \rangle \bigr\rangle
  =
  0.
  \end{aligned}
  $$
  Next, for $t = t'$, we have
  $$
  \zeta_i(t)
  = h_i(t) - \langle h_i(t) \rangle
  = \delta_{i, i(t)} - p_i,
  $$
  and
  $$
  \begin{aligned}
  \left\langle
    \zeta_i(t) \, \zeta_j(t)
  \right\rangle
  &=
  \left\langle
    \delta_{i, i(t)} \, \delta_{j, i(t)}
  \right\rangle
  -
  p_i \, p_j
  \\
  &=
  \left\langle
    \delta_{i, i(t)}
  \right\rangle
  \delta_{i, j}
  -
  p_i \, p_j
  =
  p_i \, \delta_{ij}
  - p_i \, p_j.
  \end{aligned}
  $$
  %
  The average
  $\left\langle \delta_{i, i(t)} \right\rangle$ is $p_i$,
  because as $i(t)$ goes through $1$ to $n$,
  $\delta_{i, i(t)}$ gives $1$
  if $i(t)$ matches $i$ with probability $p_i$,
  or $0$ otherwise.
}
%
Thus,
for $k \ge 2$, we get
%
\begin{equation}
  \kappa_k(t - t')
  =
  \sum_{i,j = 1}^n
  \frac{ \phi_{ki} } { p_i }
  \frac{ \phi_{kj} } { p_j }
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  %
  = \delta_{t, t'},
\notag
%\label{eq:kappa_perfect}
\end{equation}
%
where we have used
Eqs. \eqref{eq:eig_orthonormal_rows}-\eqref{eq:ortho1},
\eqref{eq:eta_def},
and
\eqref{eq:zz_perfect}.
%
\note{Proof.%Eq. \eqref{eq:kappa_perfect},
  $$
  \begin{aligned}
  \kappa_k(t - t')
  &=
  \sum_{i,j = 1}^n
  \frac{ \phi_{ki} } { p_i }
  \frac{ \phi_{kj} } { p_j }
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  \\
  %
  &=
  \sum_{i,j = 1}^n
  \frac{ \phi_{ki} } { p_i }
  \frac{ \phi_{kj} } { p_j }
  \left(
    p_i \, \delta_{i, j} - p_i \, p_j
  \right)
  \, \delta_{t, t'}
  \\
  &=
  \left(
    \sum_{j = 1}^n
    \frac{ \phi_{kj}^2 } { p_j }
    -
    \sum_{i = 1}^n \phi_{ki}
    \sum_{j = 1}^n \phi_{kj}
  \right)
  \, \delta_{t, t'}
  \\
  &=
  (1 - 0 \cdot 0) \, \delta_{t, t'}.
  \end{aligned}
  $$
  where,
  we have used
  Eq. \eqref{eq:eta_def}
  on the first line,
  %
  Eq. \eqref{eq:zz_perfect}
  on the second line,
  and
  Eq. \eqref{eq:eig_orthonormal_rows}
  on the last line.
}

\subsection{\label{sec:Gamma_onestep}
One-step process}

Another special case is the
one-dimensional one-step process\cite{vankampen},
which models a local sampling process.
In each step, the system hops to
one of the two neighboring bins
with equal probability $1/2$.
%
The transition matrix is
%
\begin{equation}
  \arraycolsep=3.6pt\def\arraystretch{1.4}
  \mathbf A
  =
  \left(
    \begin{array}{cccccccc}
      \frac \sigma 2 & \frac 1 2 & 0 & \dots & \frac {1 - \sigma} 2 \\
      \frac 1 2 & 0         & \frac 1 2 & \dots & 0 \\
      \vdots & &  & & \vdots \\
      \frac {1 - \sigma} 2 & \dots &  & \frac 1 2 & \frac \sigma 2
    \end{array}
  \right),
\notag
%\label{eq:T_nn}
\end{equation}
%
where $\sigma = 0$ for a periodic variable, or $1$ otherwise.
%
For many updating schemes
(cf. Sec. \ref{sec:band-matrix}),
we can further assume that
the matrices $\mathbf A$ and $\mathbf w$
share the eigenvectors, but not the eigenvalues.
%
The eigenvalues of $\mathbf A$ are
$$
\Lambda_k
=
1 -
2 \sin^2 \frac{ (k - 1) \, g \, \pi }
              {       2 \, n        }
,
$$
where
$g = 2$ for a periodic variable, or $1$ otherwise.
%Note that these $\Lambda_k$'s are
%not to be confused with the eigenvalues of $\mathbf w$,
%$\lambda_k$'s.
%
In this case,
the autocorrelation function $\kappa_k(t)$
decays as $\kappa_k(0) \, \Lambda_k^t$,
with $\kappa_k(0) = 1$ being the same as
the value in perfect sampling.
%
Thus,
%$\Gamma_k$ is roughly twice the autocorrelation time, and
from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
  \Gamma_k
  =
  1 + 2 \, \sum_{t = 1}^\infty \Lambda_k^t
  =
  \frac{ 1 + \Lambda_k }
       { 1 - \Lambda_k }
  =
  \cot^2 \left[
    \frac{ (k - 1) \, g \, \pi }
         { 2 \, n }
  \right]
  .
\label{eq:Gamma_onestep}
\end{equation}

\subsection{\label{sec:Gamma_measure}
Measurement via an adaptive FDS simulation
}

%To use Eq. \eqref{eq:q_opt} in a real simulation,
%we need to know both $\lambda_k$'s and $\Gamma_k$'s.
%
We can also
use Eq. \eqref{eq:y2_eql}.
measure $\Gamma_k$'s in
a simulation with a constant schedule,
$\alpha(t) = \alpha_0$.
%
%A practical difficulty is that
However,
we often do not know the intrinsic distribution, $p^*_i$,
in advance to compute
the shifted bias potential, $v_i$,
from the original one, $V_i$
(cf. Sec. \ref{sec:FDS}).
%
To circumvent the problem, we
modify Eq. \eqref{eq:y2_eql} as
%
\begin{equation}
  \operatorname{var} Y_k
  =
  \frac{1}{2}
  \alpha \, \Gamma_k \, \lambda_k,
\label{eq:varY}
\end{equation}
%
where
``$\operatorname{var}$''
means the variance from time averaging,
and
%
\begin{align}
Y_k
&=
\sum_{ i = 1 }^n
  \phi_{k i} \, (V_i - \bar V),
\label{eq:Y_def}
\end{align}
%
with
$
\bar V
\equiv
\sum_{ j = 1 }^n p_j \, V_j.
$

@@

Here we give a program of computing
the optimal schedule and the error
of a given adaptive FDS simulation.
%
Basically, we need to extract
$\lambda_i$'s and $\Gamma_i$'s
from the input,
and then feed them to Eq. \eqref{eq:q_opt}
for the schedule,
as shown in Fig. \ref{fig:vardep}.
%
%The dependence of variables are shown in Fig. \ref{fig:vardep}.
%
While the updating scheme alone
determines $\lambda_k$'s,
$\Gamma_k$'s depend also on the system
and the underlying sampling process (MC or MD),
and need to be measured using the technique
in Sec. \ref{sec:Gamma_measure}.
%
The complete program is the following.

@@
